Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-01-11 PAPER.tex
made edits from josh until 3.1.4

For our analysis, we use $8.1$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (A15 uses a slightly different RA 
range of $0$-$8.6$ hours). All power spectrum results are produced for a center frequency of $151$ MHz using a width of $10$ 
MHz ($20$ channels), identical to the analysis in A15. We note that, besides using only one baseline type instead of the three in 
A15, the PAPER-64 dataset that we use in this case study differs from that in A15 mainly by the applied fringe-rate filter. In A15, the 
applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent 
modes and reduce signal loss, though as we will explain in the next section, this signal loss was still under-estimated. With the development of a new, 
robust method for assessing signal loss, we choose to use the optimal filter in order to maximize sensitivity. This filter is 
computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate 
domain and time domain is shown in Figure \ref{fig:frp}.

@@

\subsubsection{Signal Loss Methodology} 
\label{sec:siglossmethod}
To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ 
cm signal visibilities. Here we use the same method as in A15, where mock Gaussian noise visibilities (mock EoR signals) 
are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.\footnote{One 
specific change from A15 is that we add this simulated signal into the analysis pipeline before the final fringe-rate filter is 
applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased 
estimate of signal loss (by a factor of $\sim$$10$), likely due to the use of the fringe-rate filter as a simulator. However, this pipeline difference, while significant, is not the dominant reason why signal loss was under-estimated in A15 (the dominant reason is explained in the main text).}

@@

\begin{equation}
\label{eq:Pin}
P_{\rm in,\alpha} \equiv M_{\alpha}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e}
\end{equation}

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{\rm out,\alpha} &\equiv& \widehat{\textbf{p}}_{r,\alpha}-\widehat{\textbf{p}}_{x,\alpha} \nonumber \\
&=& M_{\alpha}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - M_{\alpha}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where, for illustrative purposes and notational simplicity, we have written these equations with diagonal normalization matrices $M$, even though for our numerical results we use a full non-diagonal matrix $\mathbf{M}$ as in Equation \eqref{eq:phat}. $P_{\rm in}$ represents the unweighted power spectrum of $\textbf{e}$, our simulated EoR signal, 
$\widehat{\textbf{p}}_{x}$ is the power spectrum of the data alone, $\widehat{\textbf{p}}_{r}$ is the power spectrum of the
data plus injection, and $P_{\rm out}$ is the difference between $\widehat{\textbf{p}}_{r}$ and $\widehat{\textbf{p}}_{x}$.

In short, we can summarize a signal loss estimate as the ratio of $P_{\rm out}/P_{\rm in}$, evaluated at the data level $\widehat{\textbf{p}}_{x}$. We motivate the fact that we can evaluate the output-to-input power spectrum ratio at $\widehat{\textbf{p}}_{x}$ by the following reasoning.

In a limit with no instrumental noise, the data that we measure, $\x$, is comprised of two signals, such that
\begin{equation}
\x \equiv \f + \s,
\end{equation}
where $\mathbf{f}$ represents the foregrounds and $\mathbf{s}$ represents the cosmological signal. In general, suppose that our power spectrum algorithm passes the data through some function $g$, yielding a lossy estimate of the power spectrum $\phat_{x}$. This can be parametrized as
\begin{equation}
\label{eq:LinearPspecSum}
\langle \phat_{x} \rangle  = \langle g(\x) \rangle = \ell_{\rm fg} \p_{\rm fg} + \ell_{\rm eor} \p_{\rm eor},
\end{equation}
where $\ell_{\rm eor}$ and $\ell_{\rm fg}$ are multiplicative factors accounting for the signal loss in the true EoR power spectrum $\p_{\rm eor}$ and true foreground power spectrum $\p_{\rm fg}$, respectively. It is not \emph{a priori} obvious why this parameterization is suitable; we thus provide a toy model in Appendix \ref{sec:sigloss_appendix} to motivate this, although it should be noted that the result is an approximation.

@@

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &=& M_{\alpha}(\textbf{x}+\textbf{e})^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}(\textbf{x}+\textbf{e}) - 
M_{\alpha}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x} \nonumber \\
&=& M_{\alpha}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} + M_{\alpha}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& M_{\alpha}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} + M_{\alpha}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \nonumber \\
&-& M_{\alpha}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}
We see that $P_{\rm out}$ is comprised of multiple terms. For illustrative purposes in this subsection only, we build intuition for signal loss in $P_{out}$ by considering a scenario where $\textbf{e}$ dominates compared to $\textbf{x}$. It is in this limit where we expect signal loss to be greatest, and $P_{out}$ becomes:

\begin{eqnarray}
\label{eq:pout_expand}
P_{\rm out, \alpha} &\approx&  M_{\alpha}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} +
 2M_{\alpha}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e}.
\end{eqnarray}
The first term of $P_{\rm out}$ is simply the weighted power spectrum of $\textbf{e}$ alone. It is this quantity that was used in A15 to calculate signal loss. However, without taking into account the cross-terms (the second term), we can substantially under-estimate loss. In short, this is because correlations between the data and injected EoR signal emerge as statistical noise that is added to the power spectrum estimate, decreasing the value of $P_{\rm out}$. Including these cross-terms in our new analysis represents the bulk change of our revised analysis compared to A15.

@@

\begin{eqnarray}
P_{\rm in,\alpha} &=&  M_{\alpha}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e} \\
P_{\rm out,\alpha} &=&  M_{\alpha}\textbf{r}^{\dagger}\widehat{\textbf{C}}_{r}^{-1}\textbf{Q}^{\alpha}\widehat{\textbf{C}}_{r}^{-1}\textbf{r} \nonumber \\
&&\phantom{=====} 
-  M_{\alpha}\textbf{x}^{\dagger}\widehat{\textbf{C}}_{x}^{-1}\textbf{Q}^{\alpha}\widehat{\textbf{C}}_{x}^{-1}\textbf{x} 
\end{eqnarray}

@@

As a final complication to our procedure, we note that there will be some intrinsic scatter in our likelihood due to our having only a finite number of simulations. This is shown by the smeared `heat-map' around the solid black diagonal line (representing unity-transfer) in Figure \ref{fig:sigloss_transfercurve}. To separate the intrinsic stochasticity of signal loss from that which arises due to simulation noise, we repeat our analysis for a power spectrum estimator without signal loss ($\textbf{R} = \textbf{I}$), shown as the right plot in Figure \ref{fig:sigloss_transfercurve}. The width of the lossless scenario's signal loss likelihood is then deconvolved from the lossy scenario's signal loss likelihood. Specifically, we deconvolve Gaussian distributions of $P_{out}$ in the unweighted case from those of the weighted case (vertical cuts through the plots in Figure \ref{fig:sigloss_transfercurve}), for every $P_{in}$. By doing this, we are left with only the intrinsic scatter in signal loss, or scatter that stems from how much the random EoR signal $\textbf{e}$ happens to 
look like the data $\textbf{x}$, a quantity we do not know offhand but one that we would like to correct for. As an extreme example, if we are very unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{\rm out}$ value could arise from a range of $P_{\rm in}$ values.

@@

To conclude this method, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal 
loss estimation in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted 
power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and 
gray points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss estimation, it 
is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). 
Post-estimation, the power spectrum errors blow up to be higher than both the theory and unweighted power spectrum, a consequence of not being able to characterize covariances well using the time-averaged fringe-rate filtered data (there is less independent information to use). We elaborate on this point in the next section, as well as investigate alternate 
weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

>>>>

For our analysis, we use $8.1$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (A15 uses a slightly different RA 
range of $0$-$8.6$ hours). All power spectrum results are produced for a center frequency of $151$ MHz using a width of $10$ 
MHz ($20$ channels), identical to the analysis in A15. We note that, besides using only one baseline type instead of the three as in 
A15, the PAPER-64 dataset that we use in this case study differs from that in A15 only in the analysis. The most significant changes occur in the power spectrum analysis, which is explained in the rest of this paper, but we also note that the applied fringe-rate filter is also slightly different. In A15, the 
applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent 
modes and reduce signal loss, though as we will explain in the next section, this signal loss was still under-estimated. With the development of a new, 
robust method for assessing signal loss, we choose to use the optimal filter in order to maximize sensitivity. This filter is 
computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate 
domain and time domain is shown in Figure \ref{fig:frp}.

@@

\subsubsection{Signal Loss Methodology} 
\label{sec:siglossmethod}
To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ 
cm signal visibilities. Here we use the same method as in A15, where mock Gaussian noise visibilities (mock EoR signals) 
are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.\footnote{One 
specific change from A15 is that we add this simulated signal into the analysis pipeline before the final fringe-rate filter is 
applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased 
estimate of signal loss, %(by a factor of $\sim$$10$), 
likely due to the use of the fringe-rate filter as a simulator. However, this pipeline difference, while significant, is not the dominant reason why signal loss was under-estimated in A15 (the dominant reason is explained in the main text).}

@@

\begin{equation}
\label{eq:Pin}
P_{\rm in,\alpha} \equiv M^{\alpha}_{\rm in}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e}
\end{equation}

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{\rm out,\alpha} &\equiv& \widehat{\textbf{p}}_{r,\alpha}-\widehat{\textbf{p}}_{x,\alpha} \nonumber \\
&=& M^{\alpha}_{r}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where, for illustrative purposes and notational simplicity, we have written these equations with diagonal normalization matrices $M$, even though for our numerical results we use a full non-diagonal matrix $\mathbf{M}$ as in Equation \eqref{eq:phat}. $P_{\rm in}$ represents the unweighted power spectrum of $\textbf{e}$, our simulated EoR signal, 
$\widehat{\textbf{p}}_{x}$ is the power spectrum of the data alone, $\widehat{\textbf{p}}_{r}$ is the power spectrum of the
data plus injection, and $P_{\rm out}$ is the difference between $\widehat{\textbf{p}}_{r}$ and $\widehat{\textbf{p}}_{x}$.

In short, we can approximate a signal loss estimate as the ratio of $P_{\rm out}/P_{\rm in}$, evaluated at the data level $\widehat{\textbf{p}}_{x}$. We motivate the fact that we can evaluate the output-to-input power spectrum ratio at $\widehat{\textbf{p}}_{x}$ by the following reasoning (and then detail our signal loss estimation in practice in the next section).

In the limit of no instrumental noise, the data that we measure, $\x$, is comprised of two signals, such that
\begin{equation}
\x \equiv \f + \s,
\end{equation}
where $\mathbf{f}$ represents the foregrounds and $\mathbf{s}$ represents the cosmological signal. In general, suppose that our power spectrum algorithm passes the data through some function $g$, yielding a lossy estimate of the power spectrum $\phat_{x}$. This can be parametrized as
\begin{equation}
\label{eq:LinearPspecSum}
\langle \phat_{x} \rangle  = \langle g(\x) \rangle = \ell_{\rm fg} \p_{\rm fg} + \ell_{\rm eor} \p_{\rm eor},
\end{equation}
where $\ell_{\rm eor}$ and $\ell_{\rm fg}$ are multiplicative factors accounting for the signal loss in the true EoR power spectrum $\p_{\rm eor}$ and true foreground power spectrum $\p_{\rm fg}$, respectively. It is not \emph{a priori} obvious why this parameterization is suitable; we thus provide a toy model in Appendix \ref{sec:sigloss_appendix} to motivate this, although it should be noted that the result is an approximation.

@@

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &=& M^{\alpha}_{r}(\textbf{x}+\textbf{e})^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}(\textbf{x}+\textbf{e}) - 
M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x} \nonumber \\
&=& M^{\alpha}_{a}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} + M^{\alpha}_{b}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& M^{\alpha}_{c}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} + M^{\alpha}_{d}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \nonumber \\
&-& M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}
We see that $P_{\rm out}$ is comprised of multiple terms. For illustrative purposes in this subsection only, we build intuition for signal loss in $P_{out}$ by considering a scenario where $\textbf{e}$ dominates compared to $\textbf{x}$. It is in this limit where we expect signal loss to be greatest, and $P_{out}$ becomes:

\begin{eqnarray}
\label{eq:pout_expand}
P_{\rm out, \alpha} &\approx&  M^{\alpha}_{b}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} +
 2M^{\alpha}_{c}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e}.
\end{eqnarray}
The first term of $P_{\rm out}$ is simply the weighted power spectrum of $\textbf{e}$ alone. It is this quantity that was used in A15 to calculate signal loss. However, without taking into account the cross-terms (the second term), we can substantially under-estimate loss. In short, this is because correlations between the data and injected EoR signal emerge as statistical noise that is added to the power spectrum estimate, decreasing the value of $P_{\rm out}$. Including these cross-terms in our new analysis represents the bulk change of our revised analysis compared to A15.

@@

\begin{eqnarray}
P_{\rm in,\alpha} &=&  M^{\alpha}_{\rm in}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e} \\
P_{\rm out,\alpha} &=&  M^{\alpha}_{r}\textbf{r}^{\dagger}\widehat{\textbf{C}}_{r}^{-1}\textbf{Q}^{\alpha}\widehat{\textbf{C}}_{r}^{-1}\textbf{r} \nonumber \\
&&\phantom{=====} 
-  M^{\alpha}_{x}\textbf{x}^{\dagger}\widehat{\textbf{C}}_{x}^{-1}\textbf{Q}^{\alpha}\widehat{\textbf{C}}_{x}^{-1}\textbf{x} 
\end{eqnarray}

@@

As a final complication to our procedure, we note that there will be some intrinsic scatter in our likelihood due to our having only a finite number of simulations. To separate the intrinsic stochasticity of signal loss from that which arises due to simulation noise, we repeat our analysis for a power spectrum estimator without signal loss ($\textbf{R} = \textbf{I}$), shown as the right plot in Figure \ref{fig:sigloss_transfercurve}. This intrinsic scatter is shown as the smeared `heat-map' around the solid black diagonal line (representing unity-transfer) in the plot. The width of the lossless scenario's signal loss likelihood is then deconvolved from the lossy scenario's signal loss likelihood. Specifically, we deconvolve Gaussian distributions of $P_{out}$ in the unweighted case from those of the weighted case (vertical cuts through the plots in Figure \ref{fig:sigloss_transfercurve}), for every $P_{in}$. By doing this, we are left with only the intrinsic scatter in signal loss, or scatter that stems from how much the random EoR signal $\textbf{e}$ happens to 
look like the data $\textbf{x}$, a quantity we do not know offhand but one that we would like to correct for. As an extreme example, if we are very unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{\rm out}$ value could arise from a range of $P_{\rm in}$ values.

@@

To conclude this method, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal 
loss estimation in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted 
power spectrum $2\sigma$ upper limit, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and 
gray points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss estimation, it 
is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). 
Post-estimation, the power spectrum errors blow up to be higher than both the theory and unweighted power spectrum, a consequence of not being able to characterize covariances well using the time-averaged fringe-rate filtered data (there is less independent information to use). We elaborate on this point in the next section, as well as investigate alternate 
weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR signal.
